---
title: "Ejercicio integrador -- 09 de octubre de 2026"
format: pdf
---

This document contains the information needed for the comprehensive example for the last day of the short course on mixed models. 
Students are expected to load the data, comprehend it, propose a statistical model, and analyze it within the morning. 
Then, students will prepare a presentation on their data, model, and solution. 

## Libraries 

```{r warning=TRUE, message=FALSE}
library(tidyverse)
library(lme4)
```

## Data 

A researcher has spent their past 10 years studying the effect of nitrogen management on corn crops. 
They have collected all their data on this topic on a tidy and detailed data set, named `data_detailed` below. 
All trials were randomized complete block designs. 


Last week, they received a dataset with more data on the same topic, from a fellow researcher. 
Said researcher, however, could only recover site-level data from their dataset. All trials were randomized complete block designs. 

Their data is stored in the dataframe `data_rough` below. 

### Exercises 

Take a look at the data below and complete the following exercises:

0. Make sure you explore the data properly and are able to answer: 

- How many years does `data_detailed` contain? 
- Do all years contain all treatments? 
- Do all years contain the same number of repetitions? 
- Do `data_detailed` and `data_rough` contain the same number of observations?


1. What is the structure in the data? Please provide a drawing of the data. 

2. Write a statistical model describing the data. 

3. What is the between-year variance? What does it mean?

4. Is your model the only option to solve this problem? What are other ways someone could turn these data into results and indicate why your solution is better. 

5. Provide two other applied data examples. 

 

```{r}
set.seed(33)
data_detailed <- expand.grid(year = factor(2015:2020),
                             block = factor(1:3),
                             nitrogen = c(0,  100, 200)) |> 
  bind_rows( expand.grid(year = factor(2021:2026),
                             block = factor(1:4),
                             nitrogen = c(0, 50, 100, 150, 200)))
  
n_years <- n_distinct(data_detailed$year)
n_blocks <- n_distinct(data_detailed$year)* n_distinct(data_detailed$block)

year_eff <- rnorm(n_years, 0, 3)
block_eff <- rnorm(n_blocks, 0, 1.5)
resid <- rnorm(nrow(data_detailed), 0, 1.5)


data_detailed$mu <- ifelse(data_detailed$nitrogen<150, 
             7.3 + .062*data_detailed$nitrogen -0.00013*data_detailed$nitrogen,
             7.3 + .062*150 -0.00013*150 )

data_detailed <- data_detailed |> 
  mutate(y = mu + year_eff[year] + block_eff[block]) 

data_detailed |> 
  ggplot(aes(nitrogen, y))+
  geom_point(aes(color = year))
```


```{r}
set.seed(22)
data_rough <- expand.grid(year = factor(2006:2014),
                          block = factor(1:4),
                          nitrogen = c(0, 50, 100, 150, 200))  
  
n_years <- n_distinct(data_rough$year)
n_blocks <- n_distinct(data_rough$year)* n_distinct(data_rough$block)

year_eff <- rnorm(n_years, 0, 2.5)
block_eff <- rnorm(n_blocks, 0, 1.5)
resid <- rnorm(nrow(data_rough), 0, 1.5)


data_rough$mu <- ifelse(data_rough$nitrogen<150,  
             7.3 + .062*data_rough$nitrogen -0.00013*data_rough$nitrogen,
             7.3 + .062*150 -0.00013*150 )

data_rough <- data_rough |> 
  mutate(y_d = mu + year_eff[year] + block_eff[block]) |> 
  group_by(year, nitrogen) |> 
  mutate(y = mean(y_d)) |> 
  ungroup() 

data_rough |> 
  ggplot(aes(nitrogen, y_d))+
  geom_point(aes(color = year))+
  geom_point(aes(y = y))
```


```{r}
data_merged <- 
  bind_rows(data_detailed, 
            data_rough |> dplyr::select(-c(y_d, block)) |> unique() |> 
              mutate(block = factor(1)))
```

```{r}
m <- lmer(y ~ nitrogen + I(nitrogen^2) + (1|year/block), data = data_merged)
```

```{r}
fixef(m)
```


```{r}
m_check <- lmer(y ~ nitrogen + I(nitrogen^2) + (1|year/block),
                data =   bind_rows(data_detailed, data_rough))
```

```{r}
data_merged_alt <- 
  bind_rows(data_detailed |> group_by(year, nitrogen) |>
              mutate(y = mean(y)) |> ungroup() |>
              dplyr::select(-c(block)) |> unique(), 
            data_rough |> dplyr::select(-c(y_d, block)) |> unique())

m_alt <- lmer(y ~ nitrogen + I(nitrogen^2) + (1|year),
              data = data_merged_alt)

```



```{r}
emtrends(m, ~nitrogen, "nitrogen")
emtrends(m_alt, ~nitrogen, "nitrogen")
emtrends(m_check, ~nitrogen, "nitrogen")
```







```{r}
data_detailed |> write.csv("../data/final_A_data_detailed.csv", row.names = F)

data_rough |> write.csv("../data/final_A_data_rough.csv", row.names = F)
```





