---
title: "Ejercicio integrador -- 09 de octubre de 2026 (Genotipos)"
format: pdf
---

This document contains the information needed for the comprehensive example for the last day of the short course on applied linear mixed models. 
Students are expected to load the data, comprehend it, propose a statistical model, and analyze it within the morning portion of the course, between 9am and 12pm. 


Please complete the exercises described below and prepare a presentation to discuss your model and results in class. 


## Libraries 

```{r warning=TRUE, message=FALSE}
library(tidyverse)
library(lme4)
```

## Data 

A researcher has spent their past 4 years studying the effect of wheat alleles on the duration between emergence and flowering (in growing degree days). 
They have collected all their data on this topic on a tidy and detailed data set, named `data` below. 
Although not all trials contained the same treatments and number of blocks, they all were randomized complete block designs. 


### Exercises 

Take a look at the data below and complete the following exercises:

0. Make sure you explore the data properly and are able to answer: 

- How many years does the dataset contain? 
- Do all years contain all treatments? 
- Do all years contain the same number of repetitions? 

1. What is the structure in the data? Please provide a drawing of the data. 

2. Write a statistical model describing the data. 

3. What is the between-year variance? What does it mean?

4. Is your model the only option to solve this problem? What are other ways someone could turn these data into results and indicate why your solution is better. 

5. Mention and describe two other applied examples where this methodology would be helpful. In your presentation, develop these examples to share with the group and generate discussions. 


```{r}
# data <- read.csv("../data/final_B_data.csv")
url <- "https://raw.githubusercontent.com/jlacasa/modelos-mixtos-fauba-2026/refs/heads/main/data/final_B_data.csv"
data <- read.csv(url)

```

