---
title: "Day 2 -- Analysis of a split-plot design"
format: html
---


The data in `df` below correspond to an experiment grown in 1995-1996 at the Scottish Crop Research Institute. 

The treatment design was a 2x70 factorial, and the design structure was a split-plot design with 4 blocks, 2 whole-plot fungicide treatments, and 70 barley varieties. 

The total area was 10 rows (North/South) by 56 beds (East/West). 

```{r}
library(agridat)
data("durban.splitplot")
df <- durban.splitplot
```


```{r}
df |> 
  ggplot(aes(bed, row))+
  geom_tile(aes(fill = block))
```
```{r}
df |> 
  ggplot(aes(bed, row))+
  geom_tile(aes(fill = paste(fung, block)), color = "black")+
  coord_fixed()
```

```{r}
df |> 
  ggplot(aes(bed, row))+
  geom_tile(aes(fill = gen), color = "black")+
  coord_fixed()
```



### 2.a. 

Using math notation, write the statistical model to analyze that data. Define all model components. 

### 2.b. 

Write an ANOVA table including source of variation and degrees of freedom. Include both the treatment and the design sources of variability. 

```{r}
modelo_mal <- lm(yield ~ fung*gen , data = df )

modelo_mal$df.residual

car::Anova(modelo_mal)
```



### 2.c. 

Fit the model in (2.a.) to the data using R software. 

```{r}
modelo_bien <-
  lmer(yield ~ fung*gen + (1|block) + (1|block:fung)  , data = df)


car::Anova(modelo_bien, test.statistic = "F")
```

```{r}
emmeans(modelo_mal, ~ fung, contr = list(c(1, -1)))
emmeans(modelo_bien, ~ fung, contr = list(c(1, -1)))
```


```{r}
emmeans(modelo_bien, ~ gen:fung, contr =
          list(c(1, -1, rep(0, (68+70)))))
```

### 2.d. 

Please provide the estimated treatment means for fungicide treatment "F1" and genotype "G42" and their 95% confidence intervals.  

### 2.e. 

Please provide the estimated variance components (i.e., the $\sigma$s in (2.a.). 

```{r}
VarCorr(modelo_bien)
```


```{r}
modelo_bien <-
  lmer(yield ~ fung*gen*nitro +
         (1|block) + (1|block:fung) + (1|block:fung:gen) , data = df)


```


##split split plot design 


The field had 4 blocks.

Each block has 2 whole plots for the genotypes.

Each whole-plot had 10 sub-plots for the 5 different kinds of dust and 2 methods of application.

Each sub-plot had 2 sub-sub-plots, one for inoculated soil and the other one for uninoculated soil.


bloque > genotipo > dust:metodo > suelo inoculado/no 

```{r}
df <- agridat::goulden.splitsplit

df |> 
  ggplot(aes(col, row))+
  coord_fixed()+
  geom_tile(aes(fill = block), color = "black")
```

```{r}
df |> 
  ggplot(aes(col, row))+
  coord_fixed()+
  geom_tile(aes(fill = paste(block, gen)),
            color = "black")

```


```{r}
df |> 
  filter(block == "B1") |> 
  ggplot(aes(col, row))+
  coord_fixed()+
  geom_tile(aes(fill = paste(dust, dry)),
            color = "black")+
  geom_point(aes(color = inoc), size =4)
```


```{r}
m <- lmer(yield ~ gen*dry*dust*inoc +
            (1|block)+
            (1|block:gen) +
            (1|block:gen:dust:dry),
          data = df)

# lmer(yield ~ gen*dry*dust*inoc +
#        (1|block/gen/dust:dry),
#        data = df)
```

