---
title: "Dia 2 -- Standard Error Demonstration"
format: html
---

## Libraries 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(emmeans)
library(lme4)
```

## Data 
(from day 1)

- Field experiment studying the effect of potassium on corn yield.
- One treatment factor: Potassium fertilizer (one-way treatment structure).
- Randomized Complete Block Design with 4 repetitions (design structure).

```{r}
url <- "https://raw.githubusercontent.com/jlacasa/modelos-mixtos-fauba-2026/refs/heads/main/data/cochrancox_kfert.csv"
df <- read.csv(url)
# df <- read.csv("../data/cochrancox_kfert.csv")

df$rep <- as.factor(df$rep)
df$K2O_lbac <- as.factor(df$K2O_lbac)
```

## Model fitting 

Because this is an RCBD, the model is 

$$y_{ij} = \mu + T_i +b_j +\varepsilon_{ij}, \ \ \ \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon),$$

and there are 2 possible assumptions to make about $b_j$: 

- Fixed 
- Random 


### Assume $b_j$ is fixed: 

```{r}
m_fixed <- lm(yield ~ K2O_lbac + rep, data = df)
```

The standard error of the mean is $se(\hat{\mu_i}) = \sqrt{\frac{\sigma^2_\varepsilon}{\#blocks}}$. 

Then:
```{r}
sigma2_fixed <- sigma(m_fixed)^2
n_blocks <- n_distinct(df$rep)

(se_mui_fixed <- sqrt(sigma2_fixed/n_blocks))
```

Which is the same as the SE of the marginal means: 

```{r}
(means_fixed <- emmeans(m_fixed, ~K2O_lbac))
```

Likewise, the confidence interval comes from doing 

$$\hat{\mu_i} \pm se(\hat{\mu_i})\cdot t_{dfe}.$$

For $dfe = 8$ and $\alpha = .05$ we get

```{r}
df_fixed <- df.residual(m_fixed)
t_crit <- qt(p = 0.975, df = df_fixed)

mu_1 <- as.data.frame(means_fixed)$emmean[1]

#lower CI
mu_1 - (se_mui_fixed * t_crit)
#upper CI
mu_1 + (se_mui_fixed * t_crit)
```



### Assume $b_j$ is random: 

```{r}
m_random <- lmer(yield ~ K2O_lbac + (1|rep), data = df)
```

The standard error of the mean is $se(\hat{\mu_i}) = \sqrt{\frac{\sigma^2_\varepsilon + \sigma^2_{block}}{\#blocks}}$. 

Then:
```{r}
sigma2_random <- sigma(m_random)^2
sigma2_block <- VarCorr(m_random)$rep[1]

n_blocks <- n_distinct(df$rep)

(se_mui_random <- sqrt((sigma2_random + sigma2_block)/n_blocks))
```

Which is the same as the SE of the marginal means: 

```{r}
(means_random <- emmeans(m_random, ~K2O_lbac))
```

Likewise, the confidence interval comes from doing 

$$\hat{\mu_i} \pm se(\hat{\mu_i})\cdot t_{dfe}.$$

For $dfe = 5.57$ and $\alpha = .05$ we get

```{r}
df_random <- as.data.frame(means_random)$df[1]
t_crit <- qt(p = 0.975, df = df_random) 

mu_1 <- as.data.frame(means_random)$emmean[1]

#lower CI
mu_1 - (se_mui_random * t_crit) 
#upper CI
mu_1 + (se_mui_random * t_crit) 
```

## Now, let's take a look at the se of mean differences

The standard error of mean differences for an RCBD is 

$$se(\hat{\mu_i}-\hat{\mu_{i'}}) = \sqrt{\frac{\sigma^2_\varepsilon}{\#blocks}}$$

Thus: 

For blocks as fixed: 
```{r}
# update se(mean difference)
(se_theta_fixed <- sqrt((2*sigma2_fixed)/n_blocks))

#do one mean comparison
(contr_fixed <- emmeans(m_fixed, ~K2O_lbac,
                        contr = list(c(1, -1, 0, 0, 0)))$contr)

df_fixed <- as.data.frame(contr_fixed)$df[1]
t_crit <- qt(p = 0.975, df = df_fixed) 

theta_1 <- as.data.frame(means_fixed)$emmean[1] - as.data.frame(means_fixed)$emmean[2]

#lower CI
theta_1 - (se_theta_fixed * t_crit) 
#upper CI
theta_1 + (se_theta_fixed * t_crit) 

#t-test
tstar <- theta_1 / se_theta_fixed
pt(tstar, df = df_fixed) * 2
```

For blocks as random: 
```{r}
# update se(mean difference)
(se_theta_random <- sqrt((2*sigma2_random)/n_blocks))

#do one mean comparison
(contr_random <- emmeans(m_random, ~K2O_lbac,
                        contr = list(c(1, -1, 0, 0, 0)))$contr)

df_random <- as.data.frame(contr_random)$df[1]
t_crit <- qt(p = 0.975, df = df_random) 

theta_1 <- as.data.frame(means_random)$emmean[1] - as.data.frame(means_random)$emmean[2]

#lower CI
theta_1 - (se_theta_random * t_crit) 
#upper CI
theta_1 + (se_theta_random * t_crit) 

#t-test
tstar <- theta_1 / se_theta_random
pt(tstar, df = df_random) * 2
```
















