---
title: "Day 3 -- Standard Error Demonstration fpr IBD & recovery of interblock information"
format: html
---

## Libraries 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(emmeans)
library(lme4)
```

## Data 

- Field experiment studying the corn genotypes' yield.
- One treatment factor: Genotype (one-way treatment structure).
- Balanced incomplete block design in corn with 4 repetitions (design structure).


```{r}
dat <- agridat::cochran.bib
```

## Model fitting 

Because this is a blocked design, the model is 

$$y_{ij} = \mu + T_i +b_j +\varepsilon_{ij}, \ \ \ \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon),$$

and there are 2 possible assumptions to make about $b_j$: 

- Fixed 
- Random 


### Assume $b_j$ is fixed (not recommended for IBDs): 

```{r}
m_fixed <- lm(yield ~ gen + loc, data = dat)
```

Because this is an incomplete design, the standard error of the mean cannot be calculated with a simple fraction like before. 
We extract it manually using the model's variance-covariance matrix (`vcov`) and the linear contrast matrix ($\mathbf{L}$) representing the marginal means:

```{r}
emm_fixed <- emmeans(m_fixed, ~gen)
L <- emm_fixed@linfct

V <- vcov(m_fixed)

(se_mui_all <- sqrt(diag(L %*% V %*% t(L))))

se_mui <- se_mui_all[1]
```

Which is the same as the SE of the marginal means: 

```{r}
(means_fixed <- emmeans(m_fixed, ~gen))
```

Likewise, the confidence interval comes from doing 

$$\hat{\mu_i} \pm se(\hat{\mu_i})\cdot t_{date}.$$

For $date = 8$ and $\alpha = .05$ we get

```{r}
df_fixed <- df.residual(m_fixed)
t_crit <- qt(p = 0.975, df = df_fixed)

mu_1 <- as.data.frame(means_fixed)$emmean[1]

#lower CI
mu_1 - (se_mui * t_crit)
#upper CI
mu_1 + (se_mui * t_crit)
```



### Assume $b_j$ is random: 

```{r}
m_random <- lmer(yield ~ gen + (1|loc), data = dat)
```

Again, because this is an incomplete design, the standard error of the mean cannot be calculated with a simple fraction like before. 
We extract it manually using the model's variance-covariance matrix (`vcov`) and the linear contrast matrix ($\mathbf{L}$) representing the marginal means:

```{r}
means_random <- emmeans(m_random, ~gen)
L_rand <- means_random@linfct

V_rand <- as.matrix(vcov(m_random))

(se_mui_all_rand <- sqrt(diag(L_rand %*% V_rand %*% t(L_rand))))

se_mui_rand <- se_mui_all_rand[1]

V_random <- as.matrix(vcov(m_random))

```

Which is the same as the SE of the marginal means: 

```{r}
means_random
```

Likewise, the confidence interval comes from doing 

$$\hat{\mu_i} \pm se(\hat{\mu_i})\cdot t_{date}.$$

For $\alpha = .05$ we get

```{r}
df_random <- as.data.frame(means_random)$df[1]
t_crit <- qt(p = 0.975, df = df_random)

mu_1 <- as.data.frame(means_random)$emmean[1]

#lower CI
mu_1 - (se_mui * t_crit)
#upper CI
mu_1 + (se_mui * t_crit)

```


## Mean comparisons

Now, let's take the models above and compare the precision for mean comparisons:

```{r}
means_fixed <- emmeans(m_fixed, ~gen,
                       contr = list(c(1,-1,rep(0, 11)),
                                    c(rep(0, 11),1,-1)))
means_fixed$contrasts


means_random <- emmeans(m_random, ~gen,
                        contr = list(c(1,-1,rep(0, 11)),
                                     c(rep(0, 11),1,-1)))
means_random$contrasts
```


```{r}
L_rand <- means_random$contrasts@linfct

V_rand <- as.matrix(vcov(m_random))

(se_mui_all_rand <- sqrt(diag(L_rand %*% V_rand %*% t(L_rand))))

se_mui_rand <- se_mui_all_rand[1]
```