# Día 4 (mañana): Análisis de ensayos multi-sitios 

## Review so far  

- Study objectives may include [prediction, estimation, and attribution](https://www.tandfonline.com/doi/abs/10.1080/01621459.2020.1762613) 
- Statistical methods have different strengths and may be more (or less) useful depending on the data-method combination. 
- Statistical inference is the most rich when we focus on effect sizes and confidence intervals. 
- Mixed models are especially useful to model data that show dependence 

## Multi-location models 

Just like nested designs (e.g., split-plot designs), we can model multi-location data fairly easily.   
The data below come from a multi-environment trial testing different genotypes under different environments.
The data are unbalanced in both the number of observations per environment and the number of observations of the different genotypes.  

```{r message=FALSE, warning=FALSE}
dat <- agridat::george.wheat |> 
  mutate(env = paste(loc,year), 
         year = as.factor(year))

dat |> 
  ggplot(aes(yield))+
  geom_histogram(aes(fill = factor(year)))+
  labs(fill = "Year")+
  theme(legend.position = "bottom")+
  facet_wrap(~loc)
```


```{r eval=FALSE, include=FALSE}
agridat::aastveit.barley.covs
```

